---
title: "Lab 1 Workup"
format: 
  html: 
    code-fold: true
    toc: true
execute:
  eval: false
---

## Load in Data

You may need to install these packages

```{r}
library(tidyverse)
library(sf)
library(here)
library(tmap)

```


What do we need to change to get the data to load in the following chunk?

```{r}
sfo_trees_df<-read_csv(here("data","sfo_trees/sfo_trees.csv")) |> 
  janitor::clean_names()
```

## Part 1: Data exploration

Look at `sfo_trees_df` dataframe. How many observations are there? What does the data include? Use the console to explore the data. %>% 

What are some interesting questions you could ask about the data?

Write them below:

197618 observations
Contains data on different trees in a given area. It contains location/site data, tree data, and legal data about the tree. 

What is the significance of site order?
Is there a correlation between legal status and spcies?
Are there more of some species than others?
How many unique species are there?
Which streets have more tree density?

How many counts of each observation by legal_status? FOcus on the highest trees?

```{r}
sfo_trees_df %>% 
  group_by(legal_status) %>%
  summarize(tree_count = n()) %>%
  slice_max(tree_count, n = 5) %>%
  ggplot() +
  geom_col(aes(x=fct_reorder(legal_status, tree_count), y = tree_count)) +
  labs(y = "Tree Count", x = "Legal Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Example 2

Describe what this piece of code does. Feel free to run it to help you understand

```{r}
permitted_mta_dpw <- sfo_trees_df |> 
  filter(legal_status == 'Permitted Site' &  caretaker %in% c("MTA","DPW"))
```
This filters the sfo_trees_df data frame, creating a new data frame called "permitted_mta_dpw" that keeps all observations with a legal_status of "Permitted Site" and also with the caretaker being either "MTA" or "DPW". 


What does the `%in%` do? Write an example to explain the `%in%` operator to your neighbor using the empty code chunk below

Filters a defined column by all of the words in a list. In this case, looks at caretaker column and keeps all with "MTA" and "DPW" 

e.g. If I wanted to filter legal_status to find only Permitted Sites or Undocumented, then I would...
```{r}
sfo_trees_df %>%
  filter(legal_status %in% c("Permitted Site", "Undocumented"))

ex <- seq(1:10)

ex %in% c(1,)

```


### Example 3

Our boss tasked us with identifying oak and pine trees in SF based on legal status, location, and planting date. Wrangle the data to plot the locations (without sf borders for now) and then summarize the number of oak and pines by legal status in a separate df. Call the wrangled dataframe `oak_pine_df` and the summary `oak_pine_count`

Write pseudocode on how you would solve it

Get rid of all columns not what we want 
filter out plants that are not pine or oak
summarize pine and oak counts by legal status


Data wrangle done below:


```{r}
oak_pine_df <- sfo_trees_df %>%
  filter(str_detect(species, "Oak") | str_detect(species, "Pine")) %>% # | is indicator for "or"
  select(species, legal_status, plant_date, latitude, longitude) %>%
  mutate(type = ifelse(str_detect(species, "Oak"), "Oak", "Pine"))


```


Summary stats

```{r}
oak_pine_count<-oak_pine_df |> 
  group_by(legal_status,type) |>
  summarize(count=n())
```



Plot the locations of the trees as a scatterplot

```{r}
ggplot(oak_pine_df, aes(x = latitude, y = longitude, color = type)) +
  geom_point() +
  theme_minimal() +
  theme(axis.title = element_blank()) +
  labs(color = "Tree Type",
       caption = "Locations of Oaks and Pines in SF")

```



### Example 4

Why have a list of native trees in the data folder.

```{r}
ca_native_df<-read_csv(here('data/sfo_trees/ca_native_spp.csv'))

# Why can I use this format instead of the usual here("data","folder","file")?
```


## Part 2: Analysis and maps

We want to see if there was a difference in planting years between a native oak *Quercus agrifolia* and a native pine *Pinus radiata*. Wrangle the data and use an appropriate statistical test to see if there is a difference.



**New Function!**

How to extract the year from the plant data column when it looks like "2022-11-16"? The package `lubridate` and function `year()` pulls out year information from date formats.  See the example code below.

```{r}
lubridate::year("2022-11-16")
```

```{r}
sfo_trees_native <- sfo_trees_df %>%
  separate(species, into = c("spp_sci", "spp_common"), sep = " :: ") %>%
  select(starts_with("spp"), plant_date, legal_status, longitude, latitude) %>%
  mutate(ca_native = (spp_sci %in% ca_native_df$scientific_name))
```


```{r}
oak_pine_year_df<- sfo_trees_native %>%
  filter(spp_sci %in% c("Quercus agrifolia", "Pinus radiata")) %>%
  mutate(plant_year = lubridate::year(plant_date))
  
# input into t.test

  t.test(plant_year ~ spp_sci, data = oak_pine_year_df)
```

Use the graph below to support your results. What does each line of code do in the ggplot?

```{r}
ggplot(oak_pine_year_df)+
  geom_histogram(aes(x=plant_year),bins = 10)+
  facet_wrap(~spp_sci,ncol=1)+
  theme_minimal()
```


Plot the selected species as points with a different colors for each year and a different shape for each species

### Spatial map:

Follow along as the instructor walks through each code chunk. In week 7 we'll cover spatial manipulation in R in far more detail. This is just to show a pretty map for all the work we did.

**Step 1**

Set the coordinate reference system

```{r}
oak_pine_sf<-oak_pine_year_df |> 
  drop_na(longitude,latitude) |> 
  st_as_sf(coords=c("longitude","latitude"))

st_crs(oak_pine_sf)<-4326

ggplot(data=oak_pine_sf)+
  geom_sf(aes(color=spp_sci))+
  theme_minimal()
```


**Step 2**

Read in the shape file for San Fransico.

```{r}
sfo_map <- read_sf(here("data", "sfo_map", "tl_2017_06075_roads.shp"))

#What do you think this does?
st_transform(sfo_map, 4326)
```

Plot it to see the outlines of streets

```{r}
ggplot(data = sfo_map) +
  geom_sf()
```


We have all the ingredients. How would you plot these two together to see the trees locations in the city?

```{r}
ggplot()+
  geom_sf(data = sfo_map, size = 0.1, color = "darkgray") +
  geom_sf(data=oak_pine_sf, aes(color=spp_sci), size = 0.5)+
  theme_void() +
  labs("Oaks and Pines in SF")

```


### Interactive Maps!

We can make maps interactive in R using the tmaps package (another reason why html docs are so cool).

There are other packages that do similar things. Consider for your shiny apps!

```{r}
tmap_mode("view")

tm_shape(oak_pine_sf)+
  tm_dots(col='spp_sci')
```



WE ARE NOT DONE UNTIL WE SAVE, STAGE, COMMIT, AND PUSH!!!!!!
